Accurate spectroscopic estimates of luminosity and distance\\ for luminous red-giant stars

### 2018-07-03 paper/sp.tex

\shorttitle{spectroscopic distance estimates}
\shortauthors{eilers, hogg, rix, ness}
\watermark{kittens}

@@

\title{Accurate spectroscopic estimates of luminosity and distance for luminous red-giant stars}

@@

\begin{abstract}\noindent
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which geometric parallaxes are not precisely measured by \gaia.
Here we use the \apogee--\gaia\ overlap to train a purely linear model of
\gaia, \zmass, and \wise\ photometry and
continuum-normalized \apogee\ spectroscopy that predicts inverse-square-root luminosity
for red stars with $0<\logg<2.2$.
The inverse-square-root is used to make the noise model close to Gaussian,
so that the training can be performed with full use of uncertainties and
without any cuts on parallax or parallax signal-to-noise ratio;
the model includes an L1 regularization that zeros out the contributions of
most spectral pixels to the parallax estimates; and the training was performed
with leave-out subsamples so no star's astrometry is used even indirectly in its
spectroscopic parallax estimate.
We re-label each star in the sample
with a new spectroscopic parallax; this new parallax has an empirical uncertainty of
10\,percent, which is more precise than the \gaia\ parallax
for the vast majority of targets.
Validation with globular and open clusters shows that the spectroscopic parallax
estimates are both accurate and precise.
These distance and luminosity estimates open up great opportunities for
mapping the Milky Way disk with the next
generation of spectroscopic surveys, and especially \sdssv.
\end{abstract}

@@

\acknowledgements
This project was developed in part at the
2018 NYC Gaia Sprint, hosted by the Center for Computational Astrophysics of
the Flatiron Institute in New York City.

>>>>

\shorttitle{spectroscopic distance estimates}
\shortauthors{eilers, hogg, rix, ness}
%\watermark{kittens}

@@

\title{Accurate spectroscopic estimates of luminosity and distance\\
       for luminous red-giant stars}

@@

\begin{abstract}\noindent
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which geometric parallaxes are not precisely measured by \gaia.
Here we use the \apogee--\gaia\ overlap to train a purely linear model of
continuum-normalized \apogee\ spectroscopy
(plus \gaia, \zmass, and \wise\ photometry)
that predicts inverse-square-root luminosity for red stars with $0<\logg<2.2$.
The inverse-square-root is used to make the noise model close to Gaussian,
so that the training can be performed with full use of uncertainties and
without any cuts on parallax or parallax signal-to-noise ratio;
the model includes an L1 regularization that zeros out the contributions of
most spectral pixels to the parallax estimates; and the training was performed
with leave-out subsamples so no star's astrometry is used even indirectly in its
spectroscopic parallax estimate.
The model is flexible enough to correct for extinction by dust without any
input of explicit extinction or reddening estimates.
We re-label each star in the sample
with a new spectroscopic parallax; this new parallax has an empirical uncertainty of
10\,percent, which is more precise than the \gaia\ parallax
for the vast majority of targets.
Validation with globular and open clusters shows that the spectroscopic parallaxes
are both accurate and precise.
These distance and luminosity estimates open up great opportunities for
mapping the Milky Way disk with the next
generation of spectroscopic surveys, and especially \sdssv.
\end{abstract}

@@

\acknowledgements
This project was developed in part at the
2018 \acronym{NYC} Gaia Sprint, hosted by the Center for Computational Astrophysics of
the Flatiron Institute in New York City.
